Fast Sorting on Modern Computers

### 2005-08-31 docs/tech_report/conclusions.tex
finishded reread of conclusions also minor edits and consistancy checks in other chapters

We were able to reduce the number of cache misses considerably using a technique
adapted from LaMarca's tiled mergesort, and managed to make memory-tuned radixsort the
fastest sort in our tests. This is due to its low instruction count, few cache
misses and lack of branch mispredictions. Although we attempted to align it
using another technique from mergesort, we found this decreased its performance.

@@

Based on our tests, shellsort performs very similarly to an $O(NlogN)$ sort,
outperforming heapsort. We note that our improved version is considerably
faster, mostly due to its reduced instruction count.

We were also able to measure it's branch mispredictions, and determine a formula
to calculate them.

\section{Branch Results}

@@

Among them is the fact that bimodal branch predictors have performed better than
two-level adaptive ones in almost every case in this research. This holds across
all comparative branches in all the sorts we surveyed, and we believe that where
explicit patterns - which can be exploited by two-level adaptive predictors - do
not exist, a bimodal predictors perform better.

@@

In heapsort, we found a pattern of branch mispredictions when using a fanout in
our 8-heap, and that this pattern matched closely that of selection sort,
discussed above. We found that though our approximated results using $H_N-1$
were not identical to our actual results - due to the arrays being very small -
our actual results did become more predictable in a similar fashion.

Also in heapsort, the number of mispredictions decreases depending on how much
the heap fans out. Despite this property, the number of mispredictions using an
8-heap was greater than when using a 4-heap, due to the increased total number
of branches.

@@

Sedgewick proposed removing individual insertion sorts in quicksort with one
elegant sweep at the end. LaMarca proposed replacing the large number of
insertion sorts in order to reduce cache misses. We note that attempting to move
the pivots, as Sedgewick does, actually increases both the number of branches
and the rate of mispredictions. We conclude that LaMarca's improvement, which
slightly improves quicksorts cache performance, also slightly improves it's
branch prediction performance.

@@

We also analysed the effects of different medians on quicksort. We determine that
not using a median to choose a pivot makes the branches more predictable than
they are biased, though at a cost of a large increase in instructions (and also
significantly greater possibility of a worst-case $O(N^2)$ sort). We also saw
that the more keys considered when determining the median, the lower the number
of branches. However, this can result in an increase in the rate of branch
mispredictions. We determine that reducing entropy is not the best way of making
quicksort perform faster, and that using a median-of-three actually achieves the
higher performance.

@@

We are able to accurately predict the number of branch mispredictions in
shellsort, using our results from insertion sort. We also found that the
improvement mode to shellsort, separating the final loop from the rest, results
in a significantly faster loop, with an extra speed boost from a side-effect
skipping an iteration with $h = 2$.

\section{Comparison with LaMarca's Results}

@@

LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
misses, instruction count and cycle count. Like LaMarca, we found that a four
heap lowered our instruction count. Our level 2 misses also concurred with
LaMarca, with only one difference: memory-tuned heapsort, being an out-of-place sort,
fills the cache fully with a smaller number of keys than base heapsort, which is
in-place. LaMarca's results do not show this. Nor do they show a difference in
cache misses between these two sorts when the array does fit in the cache.
LaMarca's base heapsort has the characteristics of an out-of-place heapsort.
However, our results show that his implicit heap improvements have led to a
significantly faster heapsort.

@@

LaMarca's initial improvements to base mergesort were expected to reduce the
instruction count. We noted a reduction with each of the steps he prescribed. We
note too that each of his improvement reduce level 2 cache misses, and the shape
of our level 2 cache miss graphs exactly match his. We also note similar
characteristics: LaMarca mentions a wobble in the instruction count graphs, due to
having to copy the final merge back; we note this same wobble.

@@

@@

Our quicksort results, again, we very similar to LaMarca's. We showed an
increase is instruction count of 25\% due to multi-quicksort - very close to
LaMarca's 20\%. The shape of our instruction count graphs were the same, and
switching to a binary search in multi-quicksort reduces the instruction count.
Like him, we found that the base quicksort had very few cache misses, and that
not using Sedgewick's insertion sort improvement reduces cache misses. We found
that multi-quicksort's k-way sort changes the slope of the graph, but our result
were not conclusive as to whether this would eventually make multi-quicksort
faster than the other versions. Our results do show it being slower for all our
test cases, due to the increase in instruction count.

@@

LaMarca determines that radixsort could not be improved in the same was as the
comparison-based sorts, and that due to poor cache performance, it ran more
slowly than quicksort. We showed, however, that one of the improvements applied
to mergesort could be applied to it, and that its resulting cache performance is
sufficiently good that its low instruction count and lack of branch
mispredictions make it the fastest sort in our survey.

@@

The results also helped answer the question of which is the best sort. However,
it is important to remember that sorts have different purposes and properties,
so this shall be considered. A comparative chart of the fastest sorts is shown
in Figure \ref{All cycles}. This shows the cycles per key of each of the major
sorts and their derivatives, measured using Pentium 4 performance counters.
None of the elementary sorts or heapsorts are on the chart, as their results
are significantly worse than the other results, which makes it difficult to
accurately see them.

@@

The best stable sort is radixsort. The best comparison-based stable sort is
double-aligned tiled mergesort. The best in-place sort is quicksort. The best in-place
stable sort we examine in this report is insertion sort. The best elementary
sort is also insertion sort; as such it should be only used for very small data
sets. The best sort for very large data sets is radixsort. The fastest overall
sort was memory-tuned radixsort, though quicksort is very close.

@@

Several discoveries were made and analysed: insertion sort has one misprediction
per key; selection sort has $O(H_N-1)$ branch mispredictions per key, which
applies to long arrays, though short arrays become more predictable the longer
they get, as in heapsort; tiling mergesort can be done in two steps, reducing
the cache miss count; aligning mergesort for a level 2 cache alone is
counter-productive - aligning it for level 1 and level 2 is significantly
faster; it is possible to avoid using a sentinel and still remove the bounds
check from a standard 2-heap, simply by reordering the steps involved;
sequential searches are far cheaper than binary searches across small lists, due
to a low branch misprediction rate. Bubblesort has terrible branch performance
in the middle, due to being partially sorted - it performs much worse than
selection sort as a result; multi-mergesort contains regular access patterns
which can be exploited by a two-level adaptive branch predictor; in almost all
other cases, two-level branch predictors are measurably less accurate than
simple bimodal predictors at predicting branches in sorting algorithms. A
median-of-3 quicksort outperforms other medians; radix sort has no branch
mispredictions; shellsort has easily calculable branch mispredictions, and
performs very similarly to an $O(NlogN)$ so.

In addition, we provide an analysis of most major sorts currently used, a
comparison of $O(NlogN)$ sorts, of different medians for quicksort, of $O(N^2)$
sorts and of cache conscious sorts. We present branch prediction performance
figures for most major sorts, and adapt cache-conscious strategies from
mergesort to make an improved radixsort the fastest performing sort in our
tests, and to make mergesort itself significantly faster.

>>>>

We were able to reduce the number of cache misses considerably using a technique
adapted from LaMarca's tiled mergesort, and managed to make memory-tuned radixsort the
fastest sort in our tests. This is due to its low instruction count, few cache
misses and lack of any branch mispredictions. Although we attempted to align it
using another technique from mergesort, we found this decreased its performance.

@@

Based on our tests, shellsort performs very similarly to an $O(NlogN)$ sort,
outperforming heapsort. We note that our improved version is considerably
faster, mostly due to its reduced instruction count. We also measured its branch
mispredictions, and determined a formula to accurately calculate them.

The improvement made to shellsort, separating the final loop from the rest, also
results in a significantly faster loop, with an extra speed boost from a
side-effect skipping an iteration with $h = 2$.

\section{Branch Prediction Results}

@@

Among them is the fact that bimodal branch predictors have performed better than
two-level adaptive ones in almost every case in this research. This holds across
all comparative branches in all the sorts we surveyed, and we believe that where
explicit patterns - which can be exploited by two-level adaptive predictors - do
not exist, bimodal predictors perform better.

@@

In heapsort, we found a pattern of branch mispredictions when using a fanout in
our 8-heap, and that this pattern matched closely that of selection sort,
discussed above. We found that though our approximated results using $H_N-1$
were not identical to our actual results - due to the arrays being very small -
our results showed the branches becoming more predictable in a similar fashion.

The result of this is that the number of mispredictions decreases depending on
how much the heap fans out. Despite this property, the number of mispredictions
using an 8-heap was greater than when using a 4-heap, due to the increased total
number of branches.

@@

Sedgewick proposed replacing individual insertion sorts in quicksort with one
elegant sweep at the end. LaMarca proposed replacing this with the previous
large number of insertion sorts in order to reduce cache misses. We note that
attempting to move the pivots, as Sedgewick does, actually increases both the
number of branches and the rate of mispredictions. We conclude that LaMarca's
improvement, which slightly improves quicksorts cache performance, also slightly
improves it's branch prediction performance.

@@

We also analysed the effects of different medians on quicksort. We determine that
not using a median to choose a pivot makes the branches more predictable than
they are biased, though at a cost of a large increase in instructions (and also
significantly greater possibility of a worst-case $O(N^2)$ sort). We also saw
that the more keys considered when determining the median, the lower the number
of branches. However, this results in an increase not only in the rate of branch
mispredictions, but also in the number of actual branch misprediction. We
determine that reducing entropy is not the best way of making quicksort perform
faster, and that using a median-of-three actually achieves the higher
performance.

@@

We are able to accurately predict the number of branch mispredictions in
shellsort, using our results from insertion sort.

\section{Cache Results and Comparison with LaMarca's Results}

In several cases, we found that direct-mapped caches performed better than
fully-associative ones, resulting in fewer misses. This was due to they're
replacement policy, and that the sorts in question were heavily optimized for
the direct-mapped caches. None of the changes slowed the fully-associatvive
caches however.

@@

LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
misses, instruction count and cycle count. Like LaMarca, we found that a 4-heap
lowered our instruction count, and we found better results by using an 8-heap
instead, as an 8-heap filled our cache lines exactly. Our level 2 misses also
concurred with LaMarca, with only one difference: memory-tuned heapsort, being
an out-of-place sort, fills the cache fully with a smaller number of keys than
base heapsort, which is in-place. LaMarca's results do not show this. Nor do
they show a difference in cache misses between these two sorts when the array
does fit in the cache. LaMarca's base heapsort has the characteristics of an
out-of-place heapsort. However, our results show that his implicit heap
improvements have led to a significantly faster heapsort.

@@

LaMarca's initial improvements to base mergesort were expected to reduce the
instruction count. We noted such a reduction with each of the steps he
proscribed. We note too that each of his improvement reduce level 2 cache
misses, and the shape of our level 2 cache miss graphs exactly match his. Our
graphs also have similar characteristics: LaMarca mentions a wobble in the
instruction count graphs, due to having to copy the final merge back; we note
this same wobble.

@@

In addition to the double alignement, we also added two new improvements. The
mergesorts were tiled for the level 1 caches as well as the level 2, slightly
reducing the number of level 1 cache misses, and we removed the need for a step
to copy the array back from the auxilliary array; instead, we presort a step
further in cases where this occurs, and find the merge naturally ends in the
correct destination.

@@

Our quicksort results, again, we very similar to LaMarca's. We showed an
increase is instruction count of 25\% due to multi-quicksort - very close to
LaMarca's 20\%. The shape of our instruction count graphs were the same, and
switching to a binary search in multi-quicksort reduced the instruction count.
Like LaMarca, we found that the base quicksort had very few cache misses, and
that not using Sedgewick's insertion sort improvement reduces cache misses. We
found that multi-quicksort's k-way sort changed the slope of the graph, but our
result were not conclusive as to whether this would eventually make
multi-quicksort faster than the other versions. Our results do show it being
slower for all our test cases, due to the increase in instruction count.

@@

LaMarca determines that radixsort could not be improved in the same was as the
comparison-based sorts, and that due to poor cache performance, it ran more
slowly than quicksort. We showed, however, that one of the improvements applied
to mergesort could be applied to it, and that its resulting cache performance is
sufficiently good that - partially due to its low instruction count and lack of
branch mispredictions - it is the fastest sort in our survey.

@@

The results also helped answer the question of which is the best sort. However,
it is important to remember that sorts have different purposes and properties,
so this shall be considered. A comparative chart of the fastest sorts is shown
in Figure \vref{All cycles}. This shows the cycles per key of each of the major
sorts and their derivatives, measured using Pentium 4 performance counters.
None of the elementary sorts or heapsorts are on the chart, as their results
are significantly worse than the other results, which makes it difficult to
accurately see them.

@@

The best stable sort is memory-tuned radixsort. The best comparison-based stable
sort is double-aligned tiled mergesort. The best in-place sort is base
quicksort. The best in-place stable sort we examine in this report is insertion
sort. The best elementary sort is also insertion sort; as such it should be only
used for very small data sets. The best sort for very large data sets is
memory-tuned radixsort. The fastest overall sort was memory-tuned radixsort,
though all the base and memory-tuned quicksort are both very close.

@@

Several discoveries were made and analysed: insertion sort has one misprediction
per key; selection sort has $O(H_N-1)$ branch mispredictions per key, which
applies to long arrays, though short arrays become more predictable the longer
they get, as in heapsort; tiling mergesort can be done in two steps, reducing
the cache miss count; aligning mergesort for a level 2 cache alone is
counter-productive - aligning it for level 1 and level 2 is significantly
faster; it is possible to avoid using a sentinel and still remove the bounds
check from a standard 2-heap, simply by reordering the steps involved;
sequential searches are far cheaper than binary searches across small lists, due
to a low branch misprediction rate. Bubblesort has terrible branch performance
in the middle, due to being partially sorted - it performs much worse than
selection sort as a result; multi-mergesort contains regular access patterns
which can be exploited by a two-level adaptive branch predictor; in almost all
other cases, two-level branch predictors are measurably less accurate than
simple bimodal predictors at predicting branches in sorting algorithms. A
median-of-three quicksort outperforms other medians; radixsort has no branch
mispredictions; shellsort has easily calculable branch mispredictions, and
performs very similarly to an $O(NlogN)$ sort.

In addition, we provide an analysis of most major sorts currently used, a
comparison of $O(NlogN)$ sorts, of different medians for quicksort, of $O(N^2)$
sorts and of cache-conscious sorts. We present branch prediction performance
figures for most major sorts, and adapt cache-conscious strategies from
mergesort to not only significantly increase the performance of mergesort, but
also to make memory-tuned radixsort the fastest performing sort in our tests.

### 2005-08-31 docs/tech_report/heap.tex
davids fixes to heap sort due to incorrect taken vs not-taken. also, first corrections to conclusions
finishded reread of conclusions also minor edits and consistancy checks in other chapters

An additional consideration is that the base heapsort has more instructions than
it's out-of-place equivalent. This means that the 4-heap will have a better
instruction count, as well as better cache performance than the base heapsort.
The 8-heap should have the best cache performance of them all, due to fully
using each cache line, taking advantage of the spatial locality of the cache
line.

@@

Figure \ref{base and memory-tuned heapsorts-1} shows the behaviour of each
branch in the base heapsort algorithm. The `\n{Creation, Has Children}' branch
refers to the check as to whether a key has children, during the creation of the
heap. The `\n{Creation, Child Comparison}' branch is the comparison between the
two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\n{Destruction, Has
Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction, Child
Promotion}'.

@@

An interesting question is whether fully unrolling this loop, and thus
separating the comparison branch into seven separate branches, has an impact on
branch prediction. Essentially, there are two effects at work if we replace the
sequence of branches with a single branch. First, the single branch predictor
will spend most of its time in the state where it predicts not-taken, rather
than taken. This is because all branches apart from the first are biased in the
taken direction. This should improve accuracy. On the other hand, on those
occasions when the bi-modal branch predictor is pushed towards predicting taken,
the next branch is highly likely to be mispredicted. This is especially true
because the first two or three branches are most likely to push the predictor in
the taken direction, whereas the later branches are those that are mostly likely
to result in a misprediction in this state (because they are almost always
not-taken).

>>>>

An additional consideration is that the base heapsort has more instructions than
its out-of-place equivalent. This means that the 4-heap will have a better
instruction count, as well as better cache performance than the base heapsort.
The 8-heap should have the best cache performance of them all, due to fully
using each cache line, taking advantage of the spatial locality of the cache
line.

@@

Figure \ref{base and memory-tuned heapsorts-1} shows the behaviour of each
branch in the base heapsort algorithm. The `\n{Creation, Has Children}' branch
refers to the check as to whether a key has children, during the creation of the
heap. The `\n{Creation, Child Comparison}' branch is the comparison between the
two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
checking if the child is larger than its parent. The same checks are later made
during the destruction of the heap, and are referred to as `\n{Destruction, Has
Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction, Child
Promotion}'.

@@

An interesting question is whether fully unrolling this loop, and thus
separating the comparison branch into seven separate branches, has an impact on
branch prediction. Essentially, there are two effects at work if we replace the
sequence of branches with a single branch. First, the single branch predictor
will spend most of its time in the state where it predicts not-taken, rather
than taken. This is because all branches apart from the first are biased in the
not-taken direction. This should improve accuracy. On the other hand, on those
occasions when the bi-modal branch predictor is pushed towards predicting taken,
the next branch is highly likely to be mispredicted. This is especially true
because the first two or three branches are most likely to push the predictor in
the taken direction, whereas the later branches are those that are mostly likely
to result in a misprediction in this state (because they are almost always
not-taken).

### 2005-08-31 docs/tech_report/report.tex
minor edits and merged in davids writing
fixed stupid mistake accidentally committed. added axis labels to bubblesort graphic
so close to being done i can taste it
finishded reread of conclusions also minor edits and consistancy checks in other chapters

%\n{x} is used for
% - proper names: Atom, SimpleScalar, Papi, GnuPlot, ValGrind, glibc
% - technical names: unsigned integer, less-than, greater-than, in-place, out-of
%   place, Abstract Data Type, records, cache hierarchy, latency, hit, miss,
%   compulsory, cold-start, capacity, conflict, cache line, cache block, offset,
%   tag, virtaul addresses, bus address, temporal locality, locality, spatial
%   locality, taken, not taken, flow control branch, comparaitve branch, static,
%   semi-static, dynamic, 1-bit, bimodal, two-level adaptive, heap, children,
%   implicit heap, heapifying, sifting, fix-up, fix-down, sift-up, sift-down,
%   Repeated-adds, Remove-min, d-heap, fanout, natural, straight, unrolled,
%   Memory management unit, tiling, k-way merge, trashing divide and conquer,
%   multi-way partition, comparision based, counting based, increments
% - web addresses: http://www.gnuplot.com, http://valgrind.kde.org
% - sort names, when highlighted: algorithm S, algorithm N
% - branch predictor names: 'Creation, child promotion', etc
\newcommand{\n}[1]{\textit{#1}}

@@

\begin{titlepage}
       \begin{center}
				\vspace*{40mm}
                \Huge Sorting in the Presence of Branch Prediction and Caches\\
                \vspace{1em}
                \Large Fast Sorting on Modern Computers\\
                \vspace{2em}
                \Large Paul Biggar \hspace{1cm} David Gregg\\
				\vspace{2em}
				\large Technical Report TR-CS-2005-XX\\
                \large Department of Computer Science,\\
				\large University of Dublin, Trinity College,\\
				\large Dublin 2, Ireland.
        \end{center}
\end{titlepage}

@@

We also developed a number of other improvements to the algorithms,
such as removing the need for a sentinel in heapsort. Overall, we
found that a cache-optimized radix sort was the fastest algorithm for
all large arrays; the absence of comparison branches means that the
algorithm causes almost no branch mispredictions.
\end{abstract}
\end{spacedpars}

>>>>

%\n{x} is used for
% - proper names: Atom, SimpleScalar, Papi, GnuPlot, ValGrind, glibc, PapiEx
% - technical names: unsigned integer, less-than, greater-than, in-place, out-of
%   place, Abstract Data Type, records, cache hierarchy, latency, hit, miss,
%   compulsory, cold-start, capacity, conflict, cache line, cache block, offset,
%   tag, virtaul addresses, bus address, temporal locality, locality, spatial
%   locality, taken, not taken, flow control branch, comparaitve branch, static,
%   semi-static, dynamic, 1-bit, bimodal, two-level adaptive, heap, children,
%   implicit heap, heapifying, sifting, fix-up, fix-down, sift-up, sift-down,
%   Repeated-adds, Remove-min, d-heap, fanout, natural, straight, unrolled,
%   Memory management unit, tiling, k-way merge, trashing divide and conquer,
%   multi-way partition, comparision based, counting based, increments
% - web addresses: http://www.gnuplot.com, http://valgrind.kde.org
% - sort names, when highlighted: algorithm S, algorithm N
% - branch predictor names: 'Creation, child promotion', etc
\newcommand{\n}[1]{\textit{#1}}

@@

\begin{titlepage}
       \begin{center}
				\vspace*{40mm}
                \Huge Sorting in the Presence of Branch Prediction and Caches\\
                \vspace{1em}
                \Large Fast Sorting on Modern Computers\\
                \vspace{2em}
                \Large Paul Biggar \hspace{1cm} David Gregg\\
				\vspace{2em}
				\large Technical Report TCD-CS-2005-57\\
                \large Department of Computer Science,\\
				\large University of Dublin, Trinity College,\\
				\large Dublin 2, Ireland.\\
				\vspace{2em}
				\large August, 2005\\
        \end{center}
\end{titlepage}

@@

We also developed a number of other improvements to the algorithms, such as
removing the need for a sentinel in heapsort. Overall, we found that a
cache-optimized radixsort was the fastest algorithm for all large arrays; the
absence of comparison branches means that the algorithm causes almost no branch
mispredictions.

% TODO add copy back
\end{abstract}
\end{spacedpars}
